---
title: "Bootstrap medians (stratified)"
author: "Anastasia Teterina"
format: pdf
execute:
  echo: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Week 10 In-class Assignment

Differences in median (with the BCa bootstrap 95% confidence intervals)\
\* maternal\
\* infant\
\* neonatal\
\* under-5\
mortality between the countries exposed to versus not exposed to armed
conflict for the year 2017

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(boot))
# filter() and lag() are functions that exist in both R packages dplyr and stats
# make default choice the functions from the dplyr package
#filter <- dplyr::filter
#lag <- dplyr::lag
```

```{r}
# rfa = ready for analysis
rfa <- read.csv(paste0(here(), "/data/final_dat_300923.csv"), stringsAsFactors = FALSE) |>
  dplyr::filter(year == 2017) |>
  select(confl.IND, mat.mort, inf.mort, neonat.mort, und5.mort)
```

```{r}
set.seed(231)

res <- matrix(NA, 4, 4)

# getting rid of NAs
rfa.mat <- rfa[!is.na(rfa[, "mat.mort"]), ]
rfa.inf <- rfa[!is.na(rfa[, "inf.mort"]), ]
rfa.neonat <- rfa[!is.na(rfa[, "neonat.mort"]), ]
rfa.und5 <- rfa[!is.na(rfa[, "und5.mort"]), ]

# maternal
getmeddiff.mat <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$mat.mort, sample_data$confl.IND, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}
bootout.mat <- boot(rfa.mat, statistic = getmeddiff.mat, strata = rfa.mat$confl.IND, R = 1000)
ci.mat <- boot.ci(boot.out = bootout.mat, conf = 0.95, type = "bca")
res[1,] <- c(nrow(rfa.mat), ci.mat$t0, ci.mat$bca[c(4,5)])

# infant
getmeddiff.inf <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$inf.mort, sample_data$confl.IND, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}
bootout.inf <- boot(rfa.inf, statistic = getmeddiff.inf, strata = rfa.inf$confl.IND, R = 1000)
ci.inf <- boot.ci(boot.out = bootout.inf, conf = 0.95, type = "bca")
res[2, ] <- c(nrow(rfa.inf), ci.inf$t0, ci.inf$bca[c(4,5)])

# neonatal
getmeddiff.neonat <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$neonat.mort, sample_data$confl.IND, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}
bootout.neonat <- boot(rfa.neonat, statistic = getmeddiff.neonat, strata = rfa.neonat$confl.IND, R = 1000)
ci.neonat <- boot.ci(boot.out = bootout.neonat, conf = 0.95, type = "bca")
res[3, ]  <- c(nrow(rfa.neonat), ci.neonat$t0, ci.neonat$bca[c(4,5)])

# under 5
getmeddiff.und5 <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$und5.mort, sample_data$confl.IND, FUN = median)
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}
bootout.und5 <- boot(rfa.und5, statistic = getmeddiff.und5, strata = rfa.und5$confl.IND, R = 1000)
ci.und5 <- boot.ci(boot.out = bootout.und5, conf = 0.95, type = "bca")
res[4, ]  <- c(nrow(rfa.und5), ci.und5$t0, ci.und5$bca[c(4,5)])


mort.names <- c("Maternal", "Infant", "Neonatal", "Under 5")
res.fin <- data.frame(mort.names, res)

# res
```

```{r}
library(knitr)
kable(res.fin, digits = c(0, 0, 1, 2, 2),
      col.names = c("Type of mortality", "N observations", "Difference", "95% CI lower", "95% CI upper"), 
      caption = "Median mortality difference between countries with vs without armed conflict in 2017")
```

Conclusion. 
For all 4 types of mortality, the estimates of median in countries with the armed conflicts is higher than in countries without armed conflicts; and, considering that in all cases 95% confidence interval doesn't include 0, we can be highly confident in this conclusion (i.e. it's highly improbable that these results are due to chance finding).
